---
title: "RLE fish analysis to determine initial and current values"
author: "Swaleh Aboud"
date: '2025-04-22'
output:
  html_document:
    code_folding: hide
  '': default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
library(tidyverse)
library(tibble)
library(ggpubr)
library(readxl)
library(kableExtra)
library(here)
library(magrittr)
```


## Loading the data
We load the density data from Kenya saved as a `csv` format
```{r message=FALSE}
#Ensure you import the data from your source directory

load(file = "../../../Exploration/Fish/Format/Tanzania_data.RDA")


# Use Kenya dataset to add Reference sites
Kenya_data<-read.csv("../../../../Kenya/Relative severity/Fish/Initial values/Kenya_data.csv")

# filter eco-region - Reference
names(Kenya_data)[names(Kenya_data)=="X...Eco_region"] <- "Eco_region"

Ref_data<- Kenya_data[Kenya_data$Eco_region == "Reference",]

# Add column to match with Tanzania data
names(Ref_data)
names(Fish_NCRA3)

names(Ref_data)[names(Ref_data)=="MPA"] <- "Management_level"
names(Fish_NCRA3)[names(Fish_NCRA3)=="sub-national"] <- "sub.national"
names(Ref_data)[names(Ref_data)=="Eco_region"] <- "Eco_region"
Ref_data$Sector<-""
Ref_data$Location <-""
Ref_data$site_corrected<-""


# Add reference site to Tanzania data
Tanzania_data <- rbind(Fish_NCRA3, Ref_data)

# Change Reference site for the management_level column from Park to NTZ

Tanzania_data$Management_level[which(Tanzania_data$Eco_region=="Reference")]<- "No Take Zone"

Tanzania_data$Management_level[which(Tanzania_data$Management_level=="No take zone")]<-"No Take Zone"


# standardize family names from old and new data
Tanzania_data$family[which(Tanzania_data$family=="Scaridae")]<-"Scarinae"
Tanzania_data$family[which(Tanzania_data$family=="Serranidae")]<-"Epinephelidae"

## It is expected that all families are surveyed in all stations
## To ensure this is the case we will fill in zeroe where the family is missing

Tanzania_data<- Tanzania_data%>%pivot_wider(names_from = family, values_from = density.ind.ha., values_fn = {mean}, values_fill = 0) %>% 
  gather(key = "family", value = "density.ind.ha.", -Eco_region, -Country, - Year, - Management_level,-Sector,-Location, -Site,-site_corrected, -`sub.national`, -Station, -Station_corrected)

# Delete Chumbe_MPA data that does not have density values - values were of Biomass

Tanzania_data<- Tanzania_data[!is.na(Tanzania_data$density.ind.ha.),]

# Mazi Bay sites - A/B/D
# add management levels - Partially protected

Tanzania_data$Management_level[Tanzania_data$Station_corrected == "MnazA" | Tanzania_data$Station_corrected == "MnazB" | Tanzania_data$Station_corrected == "MnazD" | Tanzania_data$Station_corrected == "MnazE" | Tanzania_data$Station_corrected == "MnazF"] <- "Partially restricted"

# Add Eco-region to Mnazi Bay
Tanzania_data$Eco_region[Tanzania_data$Station_corrected == "MnazA" | Tanzania_data$Station_corrected == "MnazB" | Tanzania_data$Station_corrected == "MnazD" | Tanzania_data$Station_corrected == "MnazE" | Tanzania_data$Station_corrected == "MnazF"] <- "Mnazi Bay"

Tanzania_data$sub.national[Tanzania_data$Station_corrected == "MnazA" | Tanzania_data$Station_corrected == "MnazB" | Tanzania_data$Station_corrected == "MnazD" | Tanzania_data$Station_corrected == "MnazE" | Tanzania_data$Station_corrected == "MnazF"] <- "Mnazi Bay"


```

##QAQC

```{r}
#check to make sure:
#1) Each station has the same number of years of data for each family– missing zeros 
#2) Each station only has a single year of data for each family i.e. there are no duplicates 

#the previous step to pivot_wider has helped to resolve issue 1 

#1)
Tanzania_data %>% 
  group_by(Station_corrected, family) %>% 
  summarise(n_years=n()) %>% 
  ungroup() %>% 
  group_by(Station_corrected) %>% 
  summarise(issues=diff(n_years)) %>% #find the difference between the families
  filter(issues>0) #if more than 0, there are differences

#2)
l<-Tanzania_data %>% 
  group_by(Station_corrected, family,Year) %>% 
  summarise(n_dups=n()) %>% 
  filter(n_dups>1) %>% 
  pull(Station_corrected) %>% unique()  

#identify station_corrected which have duplicates

duplicates <- Tanzania_data %>% 
  group_by(Station_corrected, family, Year, density.ind.ha.) %>% 
  filter(n() > 1) # 0



#find all those stations which have more than value for each year for each family
Tanzania_data %>% 
  filter(Station_corrected %in% l) %>% 
  select(Year, Station, Station_corrected) %>% 
  distinct() %>% 
  arrange(Station_corrected, Year) %>% 
  write.csv("Station_name_changes.csv",row.names = F)

# None



```


# Remove outliers

```{r message=FALSE}

#For groupers any sites with values > 900 indiv. Ha should be excluded due to likely observer errors.
#For parrotfish any sites with values > 4000 indiv. Ha should be excluded due to likely observer errors.
#include a 10% rule to the outlier thresholds rather than having a hard cut-off - MS suggested not to add this

# Check for possible outliers

Tanzania_outliers <- Tanzania_data[Tanzania_data$sub.national !="Reference",]

ggplot(Tanzania_outliers, aes(x = factor(sub.national), y = density.ind.ha.)) + 
  geom_boxplot()+
  geom_point()+
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  facet_wrap(~family,scales = "free_y")+
  labs(x = "sub national", y = "ind/ha")+
  theme_bw()

ggsave("outlier_check.png")

outliers <- Tanzania_data %>%
  filter(family == "Epinephelidae" & density.ind.ha. > 900 | family == "Scarinae" & density.ind.ha. > 4000)

#save
write.csv(outliers, "outlier_fish_sites.csv", row.names = F)

#remove outliers
Tanzania_data2<-Tanzania_data %>% 
    filter (family == "Epinephelidae" & density.ind.ha. <=900 | family == "Scarinae" & density.ind.ha. <=4000)

save(Tanzania_data2, file="cleaned_Tanzania_fish_data.RDA")



```


# create graph of outliers sites

```{r message=FALSE}

# Sites with outliers
#      Epinephelidae: Changu & Chumbe
#      Scarinae: Mnemba, Upangu & Kipwani

# Check for possible outliers

site_outliers <- Tanzania_data %>%
  filter(site_corrected == "Changuu" | site_corrected == "Chumbe" | site_corrected == "Mnemba" | site_corrected == "Upangu" | site_corrected == "Kipwani")

# Add outliers
site_outliers$outlier <- ""
site_outliers$outlier[which(site_outliers$family=="Epinephelidae")] <- 900
site_outliers$outlier[which(site_outliers$family=="Scarinae")] <- 4000
site_outliers$outlier<- as.numeric(as.character(site_outliers$outlier))


ggplot(site_outliers, aes(x = Year, y = density.ind.ha.)) +
  geom_point() +
  stat_smooth(se=F)+
  facet_grid(family~site_corrected, scales = "free")+
  scale_x_continuous(limits = c(1999, 2009), breaks = seq(1999, 2009, 2))+
  # Add horizontal lines from df2
  geom_hline(aes(yintercept = outlier), color = "red")+
  labs(y= "Mean density [individuals/ha]")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave("Site_outlier_trend.png")



```

## Grouping and summarising the data 
We will group  data by sites, years and fish family and summarise the data by mean density.

```{r message=FALSE}

Tanzania_ref_df <- Tanzania_data%>%
  group_by( family, Eco_region, Management_level, Year, Station_corrected)%>%
  summarise(mean_den = mean(density.ind.ha.))%>%
  group_by(family,Eco_region, Management_level)%>%
  summarise(no_sites = n_distinct(Station_corrected),
            no_years = n_distinct(Year),
    density = mean(mean_den),
    sd = sd(mean_den))

knitr::kable(Tanzania_ref_df)

```

## Stations
I have corrected some station names under the column `Station_corrected`. These are as follows

``` {r message = FALSE}
#  corrected_stations<- Tanzania_data%>% group_by(Site, Station_corrected)%>% summarise(n_corrected = n_distinct(Station))%>% filter (n_corrected>1)%>%
#   mutate(n_corrected = n_corrected-1)
# 
# knitr::kable(corrected_stations)
```

## Initial values based on protected sites and reference sites
We will work with density data since it was collected continuously.

### National initial values
Note that these summaries include density values from reference locations.
```{r message=FALSE}

initial_value <- Tanzania_data%>%filter(Management_level =="No Take Zone")%>%
  group_by( family, Year, Station_corrected)%>%
  summarise(mean_den = mean(density.ind.ha.), 
            n_sites = n_distinct(Station_corrected),
            n_years = n_distinct(Year))%>%
  group_by(family)%>%
  summarise (mean_density = mean (mean_den), sdev = sd(mean_den))

knitr::kable(initial_value)

```

### Ecoregion collapse threshold values

Note that these summaries include density values from reference locations. 
```{r message=FALSE}

Mnazi_Bay <-  Tanzania_data%>%filter(Eco_region %in% c("Mnazi Bay", "Reference"),
                                  Management_level == c("No Take Zone"))%>%
  mutate (Eco_region = "Mnazi Bay") 

Mafia_Songosongo<- Tanzania_data%>%filter(Eco_region %in% c("Mafia-Songosongo", "Reference"),Management_level == "No Take Zone")%>%
  mutate (Eco_region = "Mafia-Songosongo")

Zanzibar<- Tanzania_data%>%filter(Eco_region %in% c("Zanzibar", "Reference"),Management_level == "No Take Zone")%>%
  mutate (Eco_region = "Zanzibar")

DaresSalaam<- Tanzania_data%>%filter(Eco_region %in% c("Dar es Salaam", "Reference"),Management_level == "No Take Zone")%>%
  mutate (Eco_region = "Dar es Salaam")

Tanga<- Tanzania_data%>%filter(Eco_region %in% c("Tanga", "Reference"),Management_level == "No Take Zone")%>%
  mutate (Eco_region = "Tanga")


df<- rbind(Mnazi_Bay,Mafia_Songosongo,Zanzibar,DaresSalaam,Tanga)

  initial_value_eco <- df%>% 
  group_by(family, Eco_region,Station_corrected)%>%  #create a single site value averaging across all years
  summarise(mean_den = mean(density.ind.ha.))%>%
  ungroup() %>% 
  group_by(family, Eco_region)%>%
  summarise (no_sites = n_distinct(Station_corrected),
             # no_years = n_distinct(Year),
             initial_density = mean (mean_den),
             sdev = sd(mean_den))%>%
    mutate(collapse_threshold = ifelse(family =="Scarinae", 0.1*initial_density, 0.2*initial_density))

  knitr::kable(initial_value_eco)

```


### Initial values with collapse threshold in red
```{r message=FALSE, warning=FALSE}
ggplot(df, aes(x = Year, y = density.ind.ha.)) +
  geom_point() +
  stat_smooth()+
  scale_y_log10() + # add logarithmic scale for y-axis
  facet_grid(family~Eco_region, scales = "free")+
  # Add horizontal lines from df2
  geom_hline(data = initial_value_eco, aes(yintercept = collapse_threshold), color = "red")+
  labs(y= "Mean density [individuals/ha]")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave("Initial_value_collapse_thresholds.png")

```


### Sub-national initial values

Note that these summaries include density values from reference locations. 
```{r message=FALSE}

Mnazi_Bay <-  Tanzania_data%>%filter(sub.national %in% c("Mnazi Bay", "Reference"),
                                  Management_level != c("Open access","NA"))%>%
  mutate (sub.national = "Mnazi Bay")

Mafia_Songosongo<- Tanzania_data%>%filter(sub.national %in% c("Mafia-Songosongo", "Reference"),Management_level == "No Take Zone")%>%
  mutate (sub.national = "Mafia-Songosongo")

Zanzibar<- Tanzania_data%>%filter(sub.national %in% c("Zanzibar", "Reference"),Management_level == "No Take Zone")%>%
  mutate (sub.national = "Zanzibar")

DaresSalaam<- Tanzania_data%>%filter(sub.national %in% c("Dar es Salaam", "Reference"),Management_level == "No Take Zone")%>%
  mutate (sub.national = "Dar es Salaam")

Tanga<- Tanzania_data%>%filter(sub.national %in% c("Tanga", "Reference"),Management_level == "No Take Zone")%>%
  mutate (sub.national = "Tanga")


# kiunga<- Tanzania_data%>%filter(Eco_region %in% c("Kiunga-Lamu", "Reference"),
#                                   MPA %in% c( "Reserve", "Park"))%>%
#   mutate (Eco_region = "Kiunga-Lamu")

df2<- rbind(Mnazi_Bay,Mafia_Songosongo,Zanzibar,DaresSalaam,Tanga)

  initial_value_sub <- df2%>% group_by(family, sub.national, Station_corrected)%>%
  summarise(mean_den = mean(density.ind.ha.))%>%
  group_by(family, sub.national)%>%
  summarise (no_sites = n_distinct(Station_corrected),
             # no_years = n_distinct(Year),
             initial_density = mean (mean_den),
             sdev = sd(mean_den))
  
  knitr::kable(initial_value_sub)
  
```


Because of the very large standard deviations associated with some sub-national initial values, it makes sense to calculate a bootstrap 95% confidence interval (CI) using site/station average values for each sub-national zone and using reference values too

```{r bootstrap_CI, eval=TRUE}
# Fish density initial mean and intervals 

 station_ave<-df2%>% group_by(family, sub.national, Station_corrected)%>%
  summarise(mean_den = mean(density.ind.ha.))

g<-unique(station_ave$sub.national)


mean_fish<-station_ave %>% 
  group_by(sub.national, family) %>% 
  summarise(
    n = n(),
    mean=mean_den %>% mean() %>% round(1),
    sd=mean_den %>% sd() %>% round(1),
    IQR=mean_den %>% IQR() %>% round(1),
    min=mean_den %>% min()%>% round(1),
    max=mean_den %>% max()%>% round(1),
    cv=(sd/mean) %>% round(2)
  )


bootn=10000

boot<-data.frame(matrix(0, nrow=length(g)*2, ncol=bootn))
boot2<-data.frame(matrix(0, nrow=length(g)*2, ncol=1))

h<-station_ave$family %>% unique()

for (k in 1:2){

for (y in 1:length(g)){
  
  sample_data <- station_ave %>%
  filter(sub.national == g[y], family == h[k]) %>%
  pull(mean_den)
 
  for(i in 1:bootn){
      m <- sample(sample_data,replace=T)
     boot[y,i] <- mean(m) %>% round(1)
  }
  
  if (k==1){
 boot2$mean[y]<-boot[y,1:bootn] %>% rowMeans() %>% round(1)
 boot2$lowerCI[y]<-boot[y,1:bootn] %>% as.numeric() %>%  round(1) %>% quantile(c(.025))
 boot2$upperCI[y]<-boot[y,1:bootn] %>% as.numeric() %>%  quantile(c(.975))  %>% round(1)
 boot2$sub.national[y]<-g[y]
 boot2$family[y]<-h[k]

  } else{
 boot2$mean[y+length(g)]<-boot[y,1:bootn] %>% rowMeans() %>% round(1)
 boot2$lowerCI[y+length(g)]<-boot[y,1:bootn] %>% as.numeric() %>% quantile(c(.025)) %>%   round(1)
 boot2$upperCI[y+length(g)]<-boot[y,1:bootn] %>% as.numeric() %>% quantile(c(.975)) %>%  round(1)
 boot2$sub.national[y+length(g)]<-g[y]
 boot2$family[y+length(g)]<-h[k]
  }
}
}

mean_fish_table<-mean_fish %>% 
  left_join(boot2,by = c("sub.national", "family")) %>%
  mutate(CI=((upperCI-lowerCI)/2)%>% round(1)) %>% 
  select(sub_national=sub.national,
         family,
         n,
         mean=mean.x,
         sd,
         IQR,
         min,
         max,
         cv,
         lowerCI,
         upperCI,
         CI) %>% arrange(family)

kbl(mean_fish_table) %>% kable_classic(html_font = "Calibri") %>%  row_spec(0,bold=TRUE)

write.csv(mean_fish_table,"Criterion_D_fish_baseline.csv",row.names = F)

save(station_ave, file ="fish_sites_initial_values.RDA")

```


