---
title: "Exploring Comoros NCRA Benthic Data"
author: "Swaleh Aboud"
date: "`r Sys.Date()`"
format:
  html: 
    number-sections: true
    toc: true
    fig-cap: true
editor: visual
---

## About the Document

This document provides an exploration of the benthic dataset for Comoros, aimed at gaining a deeper understanding of the available data. It highlights key aspects of the dataset, including trends, gaps, and overall quality, while visualizing the benthic data of the country. Additionally, the document explores potential objectives and analyses that could complement the Comoros National Coral Reef Assessment (NCRA), serving as a foundational guide for further data exploration and analytical approaches.

Outlined below are the steps undertaken in this exploration process:

-   Data folders
-   Import dataset
-   Trend graphs
-   Spatial distribution of data
-   Initial vs current values Comparisons

## Data and Folders

The working directory for this project is currently hosted in a private repository to facilitate collaboration with an external contributor, Franzinho Smith. Once significant progress has been achieved, the project will be transitioned to the Organizational repository for broader access. The dataset utilized for this analysis is available and can be accessed in the CORDIO GitHub repository at the following path: `Ecological-Research-Data-Analysis/1. Data processing/Comoros/1. National_data/All_years/3. Summarise`.

```{r}
#| echo: false
#| include: false

# Setting working directory
getwd()

# Required Packages
library(dplyr)
library(magrittr)
library(ggplot2)
library(patchwork) # For combining plots
library(sf)        # For spatial data handling
library(kableExtra)
library(knitr)
library(tidyr)


```

## Datasets

-   **Initial Data**: The regional dataset was utilized due to the limited number of data points for hard coral in Comoros, with only seven points available, four of which were from Moheli. The references for these points include Quod and Bigot (2000) and Ahmada et al. (2002). No initial data was available for algae.

-   **Current Data**: This refers to the data collected from 2015 to 2024, aligned with the current values used for the Kenya and Tanzania NCRA.

-   **Complete Data:** This refers to the whole dataset for Comoros at level_1 spanning all years, with a focus on hard coral and algae groups.

-   **Comoros Shapefile**: The shapefile was obtained from Nature Earth Data, accessible via the following link: [Nature Earth Data](https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-admin-0-details/).

-   **Reef Layer Shapefile:** The reef layer data was obtained from the MEOW dataset, which was downloaded from the WWF website at the following link: [Marine Ecoregions of the World](https://www.worldwildlife.org/publications/marine-ecoregions-of-the-world-a-bioregionalization-of-coastal-and-shelf-areas)

```{r}
#| echo: false
#| include: false

# Import dataset
load("../../data_intermediate/Current_regional_ALG.RDA") # current ALG
load("../../data_intermediate/Initial_regional_ALG.RDA") # Initial ALG
load("../../data_intermediate/Current_regional_HC.RDA") # Current HC
load("../../data_intermediate/Initial_regional_HC.RDA") # Current ALG
load("../../data_intermediate/Complete_comoros_data.RDA") # Complete dataset
comoros_map <- st_read("../../data_raw/spatial/ne_10m_admin_0_map_units.shp")
comoros_region <- comoros_map %>%
  filter(NAME == "Comoros")
reef_areas <- st_read("../../data_raw/spatial/meow_ecos.shp")
```

## Trend Graphs

**Natioanl Trend**

-   The default `geom_smooth()` method was used to allow the trend line to adapt naturally to the data, ensuring accurate trend representation.
-   Algae group were provided at different names including Algae, Macroalgae and Turf algae. The category Macro-algae and Algae was used and standardised to Macro-algae

```{r}
#| echo: false
#| include: false

# Rename all ALG group to Algae
comoros_data2 <- comoros_data %>%
  mutate(level1_code = recode(level1_code,
                              "ALG"   = "Macro-algae",
                              "Algae" = "Macro-algae"))

comoros_data2$Year<- as.numeric(as.character(comoros_data2$Year))

# Filter out rows where level1_code is "Turf algae"
comoros_data2 <- comoros_data2 %>% 
  filter(level1_code != "Turf algae")

initial_alg_regional <- initial_alg_regional %>% 
  filter(`Benthic code` != "AHAL", 
         `Benthic code` != "ATRF") %>% 
  mutate(`Benthic code` = if_else(`Benthic code` == "ALG", "AMAC", `Benthic code`))


# Reorder level1_code so "Hard Coral" appears first
comoros_data2 <- comoros_data2 %>%
  mutate(level1_code = factor(level1_code, levels = c("Hard coral", "Macro-algae")))


```

```{r, fig.cap= "Fig 1: Trend in mean cover (%) over years for different benthic categories in Comoros, with raw data points and trend line (95% CI)."}
#| echo: false
#| warning: false

# Trend plot
p1 <- ggplot(comoros_data2, aes(x = factor(Year), y = mean_cover)) +
  # Points for raw data
  geom_point(shape = 21, color = "black", fill = "#69b3a2", size = 1.5, alpha = 0.7) +
  
  # Trend line with confidence interval
  geom_smooth(aes(group = 1), se = TRUE, color = "black", fill = "lightblue", alpha = 0.5) +
  
  # Facet by category (Algae, Hard Coral)
  facet_wrap(~level1_code) +
  
  # Labels and title
  labs(
    x = "Year",
    y = "Mean Cover (%)"
  ) +
  
  # Minimal theme for clarity
  theme_classic() +
  theme(
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9)
  )

p1

ggsave("trend_plot.png", plot = p1, width = 8, height = 6, dpi = 300)

```

**Sub_natioanl Trend**

-   The Grand Comore dataset contains numerous zero-cover values for Algae, particularly in the 2022 and 2023 data collected by CORDIO and the University of Comoros.

```{r, fig.cap="Fig 2: Trend in mean cover (%) over years for different benthic categories and sectors in Comoros, with raw data points and LOESS trend line (95% CI)."}
#| echo: false
#| warning: false


p2 <- ggplot(comoros_data2, aes(x = factor(Year), y = mean_cover)) +
  # Points for raw data
  geom_point(shape = 21, color = "black", fill = "#69b3a2", size = 1.5, alpha = 0.7) +
  
  # Trend line with confidence interval
  geom_smooth(aes(group = 1), method = "loess", span = 0.75, se = TRUE, color = "black", fill = "lightblue", alpha = 0.5)+
  
  # Facet by category (Algae, Hard Coral) and sub_national
  facet_grid(Sector ~ level1_code) +
  
  # Labels and title
  labs(
    x = "Year",
    y = "Mean Cover (%)",
    title = ""
  ) +
  
  # Minimal theme for clarity
  theme_classic() +
  theme(
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

p2

ggsave("trend_plot_ecoregion.png", plot = p2, width = 8, height = 6, dpi = 300)
```

## Spatial Distribution of Data

-   The majority of the reef sites surveyed were classified as partially restricted, falling under National Parks, with a few located in community reserves.

-   The Anjouan sites have a greater trend of data, with more than one year of data points, compared to the other two islands.

-   Need to investigate why some coordinates at Grand Comore are on land instead of coral reefs

```{r data-prep}
#| echo: false
#| include: false

# Check the CRS of the Comoros map and set it if missing
if (is.na(st_crs(comoros_map))) {
  st_crs(comoros_map) <- 4326  # Set CRS to WGS 84 if it's missing
}

comoros_region <- comoros_map %>%
  filter(NAME == "Comoros")


# Add missing values
current_HC$Management_level[which(current_HC$Location=="Shisiwani National Park")] <- "Partially Restricted"

current_ALG$Management_level[which(current_ALG$Location=="Shisiwani National Park")] <- "Partially Restricted"

# Summarize data for Hard Coral (current_HC)
hc_summary <- current_HC %>%
  group_by(Latitude, Longitude, Management_level) %>%
  summarize(Years = n_distinct(Year), .groups = "drop")

# Summarize data for Algae (current_ALG)
alg_summary <- current_ALG %>%
  group_by(Latitude, Longitude, Management_level) %>%
  summarize(Years = n_distinct(Year), .groups = "drop")

# Add a Category column to distinguish between Hard Coral and Algae
hc_summary$Category <- "Hard Coral"
alg_summary$Category <- "Macro-algae"

# Combine the two summarized dataframes into one
hc_alg_summary <- bind_rows(hc_summary, alg_summary)

# Remove rows with missing Latitude or Longitude
hc_alg_summary_clean <- hc_alg_summary %>%
  filter(!is.na(Latitude) & !is.na(Longitude))

hc_alg_summary_clean$Management_level[which(hc_alg_summary_clean$Management_level=="Partially Restricted")] <- "Partially restricted"



# Convert the cleaned dataframe to a spatial object
hc_alg_sf <- st_as_sf(hc_alg_summary_clean, coords = c("Longitude", "Latitude"), crs = 4326)

# Format spatial data
comoros_region <- st_transform(comoros_region, crs = 4326)
st_crs(reef_areas) <- 4326
reef_areas<- st_transform(reef_areas, 4326)
reef_areas <- st_transform(reef_areas, st_crs(comoros_region))
reef_areas_comoros <- st_intersection(reef_areas, comoros_region)


```

```{r, fig.cap="Fig 3: Spatial distribution of surveyed sites in Comoros. Points are color-coded by management level, sized by the number of years, and shaped by benthic category (Hard Coral or Algae), with separate maps for each category."}
#| echo: false

# Create the map with Hard Coral and Algae side by side with a single legend
p3 <- ggplot() +
  geom_sf(data = comoros_region, fill = "lightgray", color = "black") +  # Base map of Comoros
  geom_sf(data = reef_areas_comoros, 
          aes(fill = "Islands"),  # Add the legend entry for Reef Areas
          color = "blue", alpha = 0.4) + # Set color and transparency for reef areas
  geom_sf(data = hc_alg_sf, aes(color = Management_level, size = Years, shape = Category), alpha = 0.6) +
  facet_wrap(~Category) +  # Create side-by-side maps for Hard Coral and Algae
  scale_color_manual(values = c("Partially restricted" = "blue", 
                                "Open access" = "green")) +
  scale_size_continuous(name = "Years of Data") +  # Adjust size scale for Years of data
  scale_fill_manual(name = "Islands", values = "cyan") +  # Add a manual legend for reef areas
  # labs(title = "Spatial Distribution of Surveyed Sites") +
  theme_minimal() +
  theme(
    legend.position = "bottom", 
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.box.spacing = unit(0.5, "cm"),  # Adjust spacing between legends
    legend.justification = "center"
  ) +
  guides(shape = "none")  # Remove legend for 'Category'

p3

ggsave("map_sites_distribution.png", plot = p3, width = 8, height = 6, dpi = 300)


```

## Initial vs Current Values

-   Hard coral includes data from the current period (2015–2024) and initial values (pre-1998) separated into regional and Comoros-specific datasets.

-   Algae presents current data (2015–2024) and initial regional data (pre-1998). There is no initial data (Pre 1998) for Algae.

```{r data-prep2}
#| echo: false
#| include: false

# Prepare and select required columns for Hard Coral
hard_coral_current <- current_HC %>%
  dplyr::select(mean_cover) %>% # Select the relevant column
  rename(value = mean_cover) %>% # Rename it to 'value'
  mutate(facet = "Hard Coral", x_axis = "Current (2015-2024)") %>%
  mutate(value = as.numeric(value)) # Ensure 'value' is numeric

hard_coral_initial_regional <- initial_hc_regional %>%
  dplyr::select(HC) %>% # Select the relevant column
  rename(value = HC) %>% # Rename it to 'value'
  mutate(facet = "Hard Coral", x_axis = "Initial (Pre 1998) Regional") %>%
  mutate(value = as.numeric(value)) # Ensure 'value' is numeric

# Filter only Comoros data for Initial HC
hard_coral_comoros <- initial_hc_regional %>%
  filter(Country == "Comoros") %>% # Filter for Comoros
  dplyr::select(HC) %>%
  rename(value = HC) %>%
  mutate(facet = "Hard Coral", x_axis = "Initial (Pre 1998) Comoros") %>%
  mutate(value = as.numeric(value)) # Ensure 'value' is numeric

# Prepare and select required columns for Algae
algae_current <- current_ALG %>%
  dplyr::select(mean_cover) %>% # Select the relevant column
  rename(value = mean_cover) %>% # Rename it to 'value'
  mutate(facet = "Macro-algae", x_axis = "Current (2015-2024)") %>%
  mutate(value = as.numeric(value)) # Ensure 'value' is numeric

# Prepare and select required columns for Algae Initial (Pre 1998) Regional
algae_initial_regional <- initial_alg_regional %>%
  dplyr::select(Alg) %>% # Select the relevant column
  rename(value = Alg) %>% # Rename it to 'value'
  mutate(facet = "Macro-algae", x_axis = "Initial (Pre 1998) Regional") %>%
  mutate(value = as.numeric(value)) # Ensure 'value' is numeric

# Combine all data, including both current and initial algae data
combined_data <- bind_rows(
  hard_coral_current,
  hard_coral_initial_regional,  # Regional data for Hard Coral
  hard_coral_comoros,  # Comoros-specific HC data
  algae_current,
  algae_initial_regional # Added the Initial (Pre 1998) for Algae
)

# Reorder facets so that Hard Coral comes first
combined_data$facet <- factor(combined_data$facet, levels = c("Hard Coral", "Macro-algae"))



```

```{r, fig.cap="Fig 4: Boxplot of percentage cover across different time periods (Current, Initial Regional, Initial Comoros) and benthic categories (Algae), with jittered points to display individual data. Facets represent different categories, and colors differentiate time periods and categorie."}
#| echo: false

# Create the boxplot with individual points and fixed y-axis

p4 <- ggplot(combined_data, aes(x = x_axis, y = value, fill = x_axis)) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers from the boxplot itself
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.7) +  # Add points to show all data
  facet_wrap(~facet, scales = "fixed") +  # Fix y-axis across facets
  labs(
    title = "",
    x = "Time Period",
    y = "Percentage cover"
  ) +
  scale_fill_manual(
    values = c("Current (2015-2024)" = "skyblue", 
               "Initial (Pre 1998) Regional" = "orange", 
               "Initial (Pre 1998) Comoros" = "green", 
               "Macro-algagae" = "purple"),
    guide = "none"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", size = 12)
  )

p4

ggsave("initial_current_plot.png", plot = p4, width = 8, height = 6, dpi = 300)

```

## Summary of Baseline Data

-   Used three category of Algae (ALG, ATRF and AMAC) but not Halimeda
-   No algae baseline for Comoros; suggestions to use regional or closest region such as NE Madagascar, and N Mozambique/S Tanzania or Aldabra/Cosmoledo areas.
-   The Hard Coral baseline for Comoros, based on 7 points (4 from Moheli), suggests using the regional baseline due to the limited data and poor spatial distribution at the national level.

```{r data-prep3}
#| echo: false 
#| include: false   

# Macro-algae
initial_alg_regional$`Benthic category`[is.na(initial_alg_regional$`Benthic category`) & initial_alg_regional$`Benthic code`=="AMAC"] <- "Macro-algae" 

# Rename all ALG group to Algae 
initial_alg_regional2 <- initial_alg_regional %>%   mutate(`Benthic category` = recode(`Benthic category`, "Algae"            = "Macro-algae",
                                       "Algae macro"      = "Macro-algae",
                                       "Fleshy Algae"     = "Macro-algae",
                                       "Algal mat"        = "Macro-algae",
                              "Hard substrate with algae" = "Macro-algae"))  

initial_alg_regional2 <- subset(initial_alg_regional2,`Benthic code` != "AHAL" )  

initial_alg_regional2$Year[which(initial_alg_regional2$Year=="1987/1988")] <- "1987"  

initial_alg_regional2$Year<- as.numeric(as.character(initial_alg_regional2$Year))  

initial_hc_regional$HC<- as.numeric(as.character(initial_hc_regional$HC))  

initial_alg_regional2$Alg<- as.numeric(as.character(initial_alg_regional2$Alg)) 

```

```{r data-prep4}
#| echo: false 
#| include: false  

# Hard Coral 
mean_hc<-initial_hc_regional %>%      
  summarise(group = "Hard coral",     
            area = "regional",     
            n = n(),     
            mean=HC %>% mean() %>% round(1),     
            sd=HC %>% sd() %>% round(1),     
            IQR=HC %>% IQR() %>% round(1),     
            min=HC %>% min()%>% round(1),     
            max=HC %>% max()%>% round(1),     
            cv=(sd/mean) %>% round(2))  

mean_comoros_hc <- initial_hc_regional %>%   
  filter(Country == "Comoros") %>%  # Subset for Comoros   
  summarise(group = "Hard coral",     
            area = "Comoros",     
            n = n(),     
            mean = mean(HC, na.rm = TRUE) %>% round(1),     
            sd = sd(HC, na.rm = TRUE) %>% round(1),     
            IQR = IQR(HC, na.rm = TRUE) %>% round(1),     
            min = min(HC, na.rm = TRUE) %>% round(1),     
            max = max(HC, na.rm = TRUE) %>% round(1),     
            cv = (sd / mean) %>% round(2)   )  

# Algae 
mean_alg<-initial_alg_regional2 %>%      
  summarise(group = "Macro-algae",     
            area = "regional",     
            n = n(),     
            mean=Alg %>% mean() %>% round(1),     
            sd=Alg %>% sd() %>% round(1),     
            IQR=Alg %>% IQR() %>% round(1),     
            min=Alg %>% min()%>% round(1),     
            max=Alg %>% max()%>% round(1),     
            cv=(sd/mean) %>% round(2))  

# Macroalgae-coral ratio
# Convert Year to character in both data frames for consistent matching
initial_alg_regional2$Year <- as.character(initial_alg_regional2$Year)
initial_hc_regional$Year <- as.character(initial_hc_regional$Year)

# Ensure the key columns to match are the same type in both data frames
# You may want to convert other columns like 'data source/owner', 'Country', 'Sector', 'Site', and 'Station' to character if needed

# Match rows from both data frames based on the key columns
match_indices <- match(
  paste(initial_alg_regional2$`data source/owner`, initial_alg_regional2$Country, initial_alg_regional2$Year, 
        initial_alg_regional2$Sector, initial_alg_regional2$Site, initial_alg_regional2$Station),
  paste(initial_hc_regional$`data source/owner`, initial_hc_regional$Country, initial_hc_regional$Year,
        initial_hc_regional$Sector, initial_hc_regional$Site, initial_hc_regional$Station)
)

# Filter out rows where no match was found (NA) in the match_indices
matched_data <- initial_alg_regional2[!is.na(match_indices), ]

# Add HC column from initial_hc_regional to matched data
matched_data$HC <- initial_hc_regional$HC[match_indices[!is.na(match_indices)]]

# remove unnecessary columns like 'Benthic category' and 'Benthic code'
mean_MaCoR <- matched_data %>%
  dplyr::select(-`Benthic category`, -`Benthic code`)


mean_MaCoR <- mean_MaCoR %>%
  mutate(MaCoR = Alg / (Alg + HC)) %>%  # Create MaCoR column
  dplyr::select(-Alg, -HC)  # Remove Alg and HC columns

# Algae 
mean_macoR<-mean_MaCoR %>%      
  summarise(group = "MaCoR",     
            area = "regional",     
            n = n(),     
            mean=MaCoR %>% mean() %>% round(1),     
            sd=MaCoR %>% sd() %>% round(1),     
            IQR=MaCoR %>% IQR() %>% round(1),     
            min=MaCoR %>% min()%>% round(1),     
            max=MaCoR %>% max()%>% round(1),     
            cv=(sd/mean) %>% round(2)) 



# Merge all the three
initial_values_hc_alg <- rbind(mean_hc,mean_comoros_hc,mean_alg,mean_macoR) 

```

```{r}
#| echo: false 
#| include: false  

bootn=1000

# HC
# regional
boot_hc_regional<-data.frame(matrix(0, nrow=1, ncol=bootn))
boot2_hc_regional<-data.frame(matrix(0, nrow=1, ncol=1))

sample_hc_regional<-initial_hc_regional$HC


  for(i in 1:bootn){
      m <- sample(sample_hc_regional,replace=T)
     boot_hc_regional[1,i] <- mean(m) %>% round(1)
  }
 boot2_hc_regional$mean_bootstrap[1]<-boot_hc_regional[1,1:bootn] %>% rowMeans() %>% round(1)
 boot2_hc_regional$lowerCI[1]<-boot_hc_regional[1,1:bootn] %>% as.numeric() %>% quantile(c(.025))  %>%  round(1)
 boot2_hc_regional$upperCI[1]<-boot_hc_regional[1,1:bootn] %>% as.numeric() %>% quantile(c(.975))  %>% round(1)
 boot2_hc_regional$sd_boot[1]<-boot_hc_regional[1,1:bootn] %>% as.numeric() %>% sd() %>% round(2)
 boot2_hc_regional$area<-'regional'

 #combine bootstrapped CIs with mean and other intervals
mean_hc_regional<-mean_hc %>% 
  left_join(boot2_hc_regional, by = "area") %>%
  mutate(CI=((upperCI-lowerCI)/2)%>% round(1)) %>% 
  dplyr::select(group,     
         area,
         n,
         mean,
         sd,
         IQR,
         min,
         max,
         cv,
         lowerCI,
         upperCI,
         CI,
         sd_boot)

# Comoros
boot_hc_national<-data.frame(matrix(0, nrow=1, ncol=bootn))
boot2_hc_national<-data.frame(matrix(0, nrow=1, ncol=1))
 sample_hc_comoros<- initial_hc_regional %>% filter(Country == "Comoros") # Subset for Comoros initial_hc_regional$HC
 sample_hc_comoros<- sample_hc_comoros$HC


  for(i in 1:bootn){
      m <- sample(sample_hc_comoros,replace=T)
     boot_hc_national[1,i] <- mean(m) %>% round(1)
  }
 boot2_hc_national$mean_bootstrap[1]<-boot_hc_national[1,1:bootn] %>% rowMeans() %>% round(1)
 boot2_hc_national$lowerCI[1]<-boot_hc_national[1,1:bootn] %>% as.numeric() %>% quantile(c(.025))  %>%  round(1)
 boot2_hc_national$upperCI[1]<-boot_hc_national[1,1:bootn] %>% as.numeric() %>% quantile(c(.975))  %>% round(1)
 boot2_hc_national$sd_boot[1]<-boot_hc_national[1,1:bootn] %>% as.numeric() %>% sd() %>% round(2)
 boot2_hc_national$area<-'Comoros'

 #combine bootstrapped CIs with mean and other intervals
mean_hc_national<-mean_comoros_hc %>% 
  left_join(boot2_hc_national, by = "area") %>%
  mutate(CI=((upperCI-lowerCI)/2)%>% round(1)) %>% 
  dplyr::select(group,     
         area,
         n,
         mean,
         sd,
         IQR,
         min,
         max,
         cv,
         lowerCI,
         upperCI,
         CI,
         sd_boot)


# Macro-algae
# regional
boot_alg_national<-data.frame(matrix(0, nrow=1, ncol=bootn))
boot2_alg_national<-data.frame(matrix(0, nrow=1, ncol=1))
 sample_alg_regional<-initial_alg_regional2$Alg


  for(i in 1:bootn){
      m <- sample(sample_alg_regional,replace=T)
     boot_alg_national[1,i] <- mean(m) %>% round(1)
  }
 boot2_alg_national$mean_bootstrap[1]<-boot_alg_national[1,1:bootn] %>% rowMeans() %>% round(1)
 boot2_alg_national$lowerCI[1]<-boot_alg_national[1,1:bootn] %>% as.numeric() %>% quantile(c(.025))  %>%  round(1)
 boot2_alg_national$upperCI[1]<-boot_alg_national[1,1:bootn] %>% as.numeric() %>% quantile(c(.975))  %>% round(1)
 boot2_alg_national$sd_boot[1]<-boot_alg_national[1,1:bootn] %>% as.numeric() %>% sd() %>% round(2)
 boot2_alg_national$area<-'regional'

 #combine bootstrapped CIs with mean and other intervals
mean_alg_regional<-mean_alg %>% 
  left_join(boot2_alg_national, by = "area") %>%
  mutate(CI=((upperCI-lowerCI)/2)%>% round(1)) %>% 
  dplyr::select(group,     
         area,
         n,
         mean,
         sd,
         IQR,
         min,
         max,
         cv,
         lowerCI,
         upperCI,
         CI,
         sd_boot)

# Macroalgae to coral ratio
# regional
boot_macor_national<-data.frame(matrix(0, nrow=1, ncol=bootn))
boot2_macor_national<-data.frame(matrix(0, nrow=1, ncol=1))
 sample_macor_regional<-mean_MaCoR$MaCoR


  for(i in 1:bootn){
      m <- sample(sample_macor_regional,replace=T)
     boot_macor_national[1,i] <- mean(m) %>% round(1)
  }
 boot2_macor_national$mean_bootstrap[1]<-boot_alg_national[1,1:bootn] %>% rowMeans() %>% round(1)
 boot2_macor_national$lowerCI[1]<-boot_macor_national[1,1:bootn] %>% as.numeric() %>% quantile(c(.025))  %>%  round(1)
 boot2_macor_national$upperCI[1]<-boot_macor_national[1,1:bootn] %>% as.numeric() %>% quantile(c(.975))  %>% round(1)
 boot2_macor_national$sd_boot[1]<-boot_macor_national[1,1:bootn] %>% as.numeric() %>% sd() %>% round(2)
 boot2_macor_national$area<-'regional'

 #combine bootstrapped CIs with mean and other intervals
mean_amacor_regional<-mean_macoR %>% 
  left_join(boot2_macor_national, by = "area") %>%
  mutate(CI=((upperCI-lowerCI)/2)%>% round(1)) %>% 
  dplyr::select(group,     
         area,
         n,
         mean,
         sd,
         IQR,
         min,
         max,
         cv,
         lowerCI,
         upperCI,
         CI,
         sd_boot)




#combine HC and ALG into single table
mean_summary_table <- rbind(mean_hc_regional,mean_hc_national,mean_alg_regional,mean_amacor_regional)

```

```{r}
#| echo: false
#| warning: false 

cat("Table 1: Summary statistics for hard coral, algae and macraolgae - coral ration, including regional and Comoros-specific values (mean, SD, IQR, min, max, CV).\n\n")

kable(head(mean_summary_table)) %>% kable_classic(html_font = "Calibri")
```

## Save Data

## Baseline data

-   The output baseline file has been saved as **Criterion_D_hard_coral_baseline_comoros**.

```{r}
#| echo: false
#| include: false

# Define ecoregion values
ecoregions <- c("Anjouan Island", "Grand Comore Island", "Moheli Island","National")

# Expand dataframe by duplicating rows for each ecoregion
baseline_comoros <- mean_summary_table %>%
  tidyr::uncount(length(ecoregions)) %>%  # Duplicate rows
  mutate(ecoregion = rep(ecoregions, times = nrow(mean_summary_table))) %>%  # Add ecoregion values
  dplyr::select(ecoregion, everything())  # Reorder columns to make ecoregion the first column

# Save the data
save(baseline_comoros,file="Criterion_D_hard_coral_baseline_comoros.RDA")

```

## Current Data

```{r}
#| echo: false
#| include: false

# Hard coral
# check that for each site there is only a single benthic code per Year
h<-comoros_data %>%
  group_by(Sector, Location, Site, Year,level1_code,Organization) %>%
  summarise(n = n()) %>% 
  filter(n>1) # should be 0 if data is in order


# set current windows
current_window<-2015

#filter dataset for only HC values 

HC_current<-comoros_data %>% 
  group_by(Sector,Location,Site) %>%   
  filter(Year>=current_window, level1_code=='Hard coral')%>%
  summarise(first_year=min(Year),
            recent_year=max(Year),
            no_years=length(unique(Year)),
            recent_cover=head(mean_cover,n=1),
            current_cover=mean(mean_cover,na.rm = T),
            lat=unique(Latitude)[1],
            long=unique(Longitude)[1]) #pick just one as different coordinates creates duplicates


# Filter the rows with duplicated site_id
HC_current %>%
  group_by(Site) %>%
  filter(n() > 1) # Should be 0

# Set thresholds
HC_current$threshold_cover<-5  #change threshold cover here if required

names(HC_current)[names(HC_current)=="Sector"] <- "ecoregion"

save(HC_current, file = "Crit_D_HC_data_table.RDA")



# Macroalgae to coral ration
#filter dataset for only HC values
AMAC_current <- comoros_data %>%
  mutate(level1_code = recode(level1_code, 
                              "ALG" = "Macro-algae", 
                              "Algae" = "Macro-algae")) %>%
  filter(level1_code != "Turf algae", Year >= 2015) %>%
  dplyr::select(-sd, -se) %>%
  group_by(Year, Sector, Location, Site,Latitude,Longitude, level1_code) %>%
  summarise(mean_cover = sum(mean_cover, na.rm = TRUE), .groups = "drop") %>%
  spread(level1_code, mean_cover) %>%
  drop_na(`Hard coral`, `Macro-algae`) %>%
  mutate(macoR = `Macro-algae` / (`Macro-algae` + `Hard coral`)) %>%  # Add macoR column
  dplyr::select(-`Hard coral`, -`Macro-algae`)  # Remove Hard coral and Macro-algae columns


MaCoR_current<-AMAC_current %>% 
  group_by(Sector,Location,Site) %>%   
  summarise(first_year=min(Year),
            recent_year=max(Year),
            no_years=length(unique(Year)),
            recent_cover=head(macoR,n=1),
            current_cover=mean(macoR,na.rm = T),
            lat=unique(Latitude)[1],
            long=unique(Longitude)[1]) #pick just one as different coordinates creates duplicates


# Filter the rows with duplicated site_id
MaCoR_current %>%
  group_by(Site) %>%
  filter(n() > 1) # Should be 0

# Set thresholds
MaCoR_current$threshold_cover<-0.8333  #change threshold cover here if required

names(MaCoR_current)[names(MaCoR_current)=="Sector"] <- "ecoregion"

save(MaCoR_current, file = "Crit_D_MaCoR_data_table.RDA")

```
