---
title: "Criteria D Analysis for NCRA Comoros"
author: "Swaleh Aboud"
date: "`r Sys.Date()`"
format:
  docx: 
    number-sections: true
    toc: true
    fig-cap: true
editor: visual
---

# About the Document

# Data and Folders

The working directory for this project is currently hosted in a private repository to facilitate collaboration with an external contributor, Franzinho Smith. Once significant progress has been achieved, the project will be transitioned to the Organizational repository for broader access. The dataset utilized for this analysis is available and can be accessed in the CORDIO GitHub repository at the following path: `\Coral-Juvenile-Manuscript\Comoro NCRA\analysis_code\Benthic\Criteria D`.

```{r}
#| echo: false
#| include: false

# Setting working directory
getwd()

# Required Packages
library(dplyr)
library(magrittr)
library(ggplot2)
library(patchwork) # For combining plots
library(sf)        # For spatial data handling
library(kableExtra)
library(knitr)




```

# Datasets

```{r}
#| echo: false
#| include: false

load("../../../creation_code/Exploration/Crit_D_HC_data_table.RDA") # current data
load("../../../creation_code/Exploration/Criterion_D_hard_coral_baseline_comoros.RDA") # baseline HC
load("../../../creation_code/Exploration/Crit_D_MaCoR_data_table.RDA") # MaCoR

```

```{r}
#| echo: false
#| include: false


# Hard coral
# Baseline
baseline_HC_selected <- 
  baseline_comoros %>%
  filter(area %in% c("regional")) %>%
  filter(group %in% c("Hard coral")) %>%
  dplyr::select(ecoregion,
                group, 
                area, 
                baseline_cover = mean,
                baseline_sd = sd)


# Current
current_HC_selected <- HC_current %>%
dplyr::select(ecoregion, Location, Site, lat, long, current_cover, threshold_cover)

# Add a row with 'national' under ecoregion
current_HC_selected <- current_HC_selected %>%
  bind_rows(
    current_HC_selected %>%
      reframe(
        ecoregion = "National",
        Location = Location,
        Site = Site,
        lat = lat,
        long = long,
        current_cover = current_cover,
        threshold_cover = threshold_cover
      )
  )

current_HC_selected$ecoregion[which(current_HC_selected$ecoregion=="Grand Comore")] <-"Grand Comore Island"
current_HC_selected$ecoregion[which(current_HC_selected$ecoregion=="Moheli island")] <-"Moheli Island"


# Macroalgae to Coral ratio
# Baseline
baseline_macor_selected <- baseline_comoros %>%
  filter(area %in% c("regional")) %>%
  filter(group %in% c("MaCoR")) %>%
  dplyr::select(ecoregion,
                group, 
                area, 
                baseline_cover = mean,
                baseline_sd = sd)


# Current
current_macor_selected <- 
  MaCoR_current %>%
  dplyr::select(ecoregion, 
                Location, 
                Site, 
                lat, 
                long, 
                current_cover,
                threshold_cover)

# Add a row with 'national' under ecoregion
current_macor_selected <- current_macor_selected %>%
  bind_rows(
    current_macor_selected %>%
      reframe(
        ecoregion = "National",
        Location = Location,
        Site = Site,
        lat = lat,
        long = long,
        current_cover = current_cover,
        threshold_cover = threshold_cover
      )
  )

current_macor_selected$ecoregion[which(current_macor_selected$ecoregion=="Grand Comore")] <-"Ngazidja"
current_macor_selected$ecoregion[which(current_macor_selected$ecoregion=="Moheli island")] <-"Mwali"
current_macor_selected$ecoregion[which(current_macor_selected$ecoregion=="Anjouan Island")] <-"Ndzuani"

```

# Evaluate Criteria D

## Hard coral

```{r}
#| echo: false
#| include: false
#| warning: false

# Set iteration levels
i_min <- 10
i_max <- 1e3

# Set iteration interval
i_interval <- 10

# Create empty object to hold results
criterion_d_biotic_disruption_coral_cover_method_b <- tibble()

# Loop through iterations
for(i in seq(from = i_min, to = i_max, by = i_interval)) {

  # Set seed for reproducibility
  set.seed(i + 81)

  # Randomly assign baseline values
  dat <- current_HC_selected %>%
    mutate(
      baseline_cover = 44, # regional baseline
      baseline_sd = 18.1 # ragionel value
    ) %>%
    group_by(ecoregion) %>%
    mutate(
      baseline_cover = rnorm(1, mean = baseline_cover, sd = baseline_sd)
    )

  # Calculate relative severity
  dat <- dat %>%
    mutate(
      relative_severity = 100 * (baseline_cover - current_cover) / 
                          (baseline_cover - threshold_cover),
      relative_severity = scales::rescale(relative_severity, to = c(0, 100))
    )

  # Summarize proportion of stations for relative severity classes
  summarized <- dat %>%
    group_by(ecoregion) %>%
    summarise(
      rel_sev_30 = 100 * sum(relative_severity >= 30 & relative_severity < 50) / length(relative_severity),
      rel_sev_50 = 100 * sum(relative_severity >= 50 & relative_severity < 80) / length(relative_severity),
      rel_sev_80 = 100 * sum(relative_severity >= 80 & relative_severity <= 100) / length(relative_severity),
      relative_severity = mean(relative_severity, na.rm = TRUE), # Add mean relative severity
      .groups = "drop"
    )

  # Assign threat status
  summarized <- summarized %>%
    mutate(
      status_30 = ifelse(rel_sev_30 >= 80 & rel_sev_30 <= 100, 2, NA),
      status_50 = ifelse(rel_sev_50 >= 80 & rel_sev_50 <= 100, 3, NA),
      status_50 = ifelse(rel_sev_50 >= 50 & rel_sev_50 < 80, 2, status_50),
      status_80 = ifelse(rel_sev_80 >= 50 & rel_sev_80 < 80, 3, NA),
      status_80 = ifelse(rel_sev_80 >= 80 & rel_sev_80 <= 100, 4, status_80),
      status_80 = ifelse(rel_sev_80 >= 30 & rel_sev_80 < 50, 2, status_80),
      status_30 = ifelse(is.na(status_30), 1, status_30),
      status_50 = ifelse(is.na(status_50), 1, status_50),
      status_80 = ifelse(is.na(status_80), 1, status_80),
      max_threat = pmax(status_30, status_50, status_80)
    )

  # Create conversion object for threat values
  threat_conversions <- tribble(
    ~threat_value, ~status,
    1, "NT/LC",
    2, "VU",
    3, "EN",
    4, "CR",
    5, "CO"
  )

  # Convert threat values
  summarized <- summarized %>%
    left_join(threat_conversions %>% rename(max_threat = threat_value), by = "max_threat")

  # Include iteration number
  summarized <- summarized %>%
    mutate(Iteration = i)

  # Harvest results
  criterion_d_biotic_disruption_coral_cover_method_b <- bind_rows(
    criterion_d_biotic_disruption_coral_cover_method_b,
    summarized
  )
}


criterion_d_biotic_disruption_coral_cover_method_b$ecoregion[which(criterion_d_biotic_disruption_coral_cover_method_b$ecoregion=="Grand Comore Island")] <-"Ngazidja"
criterion_d_biotic_disruption_coral_cover_method_b$ecoregion[which(criterion_d_biotic_disruption_coral_cover_method_b$ecoregion=="Moheli Island")] <-"Mwali"
criterion_d_biotic_disruption_coral_cover_method_b$ecoregion[which(criterion_d_biotic_disruption_coral_cover_method_b$ecoregion=="Anjouan Island")] <-"Ndzuani"


summarized$ecoregion[which(summarized$ecoregion=="Grand Comore Island")] <-"Ngazidja"
summarized$ecoregion[which(summarized$ecoregion=="Moheli Island")] <-"Mwali"
summarized$ecoregion[which(summarized$ecoregion=="Anjouan Island")] <-"Ndzuani"

    
    
# summarise
    criterion_d_biotic_disruption_coral_cover_method_b %>%
      group_by(ecoregion,
               status) %>%
      summarise(n_categories = n()) %>%
      mutate(percent = 100 * n_categories / sum(n_categories))
    
 
```

```{r}
#| echo: false
#| include: false

# Create a new column for IUCN RLE categories
dat <- dat %>%
  mutate(
    iucn_status = case_when(
      relative_severity < 30 ~ "NT/LC",  # Green
      relative_severity >= 30 & relative_severity < 50 ~ "VU",  # Yellow
      relative_severity >= 50 & relative_severity < 80 ~ "EN",  # Orange
      relative_severity >= 80 & relative_severity <= 100 ~ "CR",  # Red
      TRUE ~ NA_character_
    )
  )

# Define the IUCN RLE color coding
iucn_colors <- c(
  "CR" = "red",
  "EN" = "orange",
  "VU" = "yellow",
  "NT/LC" = "green"
)

# Calculate the number of points for each ecoregion
ecoregion_counts <- dat %>%
  group_by(ecoregion) %>%
  summarise(n = n(), .groups = "drop")

dat <- dat %>%
  mutate(iucn_status = factor(iucn_status, levels = c("NT/LC", "VU", "EN", "CR")))


dat$ecoregion[which(dat$ecoregion=="Grand Comore Island")] <-"Ngazidja"
dat$ecoregion[which(dat$ecoregion=="Moheli Island")] <-"Mwali"
dat$ecoregion[which(dat$ecoregion=="Anjouan Island")] <-"Ndzuani"

ecoregion_counts$ecoregion[which(ecoregion_counts$ecoregion=="Grand Comore Island")] <-"Ngazidja"
ecoregion_counts$ecoregion[which(ecoregion_counts$ecoregion=="Moheli Island")] <-"Mwali"
ecoregion_counts$ecoregion[which(ecoregion_counts$ecoregion=="Anjouan Island")] <-"Ndzuani"

dat$ecoregion <- factor(dat$ecoregion, levels = c("Ngazidja", "Ndzuani", "Mwali", "National"))


dat_hc <- dat


```

```{r, fig.cap= "Fig. 1; Relative severity for Hard coral by ecoregion, with points color-coded according to IUCN RLE status (NT/LC, VU, EN, CR). The number of data points for each ecoregion is indicated by n"}
#| echo: false


# Create the plot
p1 <- suppressWarnings(
  ggplot(dat, aes(x = ecoregion, y = relative_severity)) +
  geom_boxplot(aes(fill = ecoregion), alpha = 0.5, outlier.shape = NA) +  # Boxplot with different fill colors for ecoregions, remove outliers from boxplot
  geom_point(aes(color = iucn_status), alpha = 0.7, size = 3, position = position_jitter(width = 0.2)) +  # Add scattered points with jitter
  scale_y_reverse(  # Reverse the Y-axis
    name = "Relative Severity",
    breaks = seq(0, 100, by = 20)
  ) +
  scale_color_manual(
    values = iucn_colors,  # Apply color mapping based on IUCN status
    name = "Relative Severity"
  ) +
  labs(
    title = "",
    x = "",
    y = "Relative Severity (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "blue", linetype = "dotted", size = 0.5),
    panel.grid.major.y = element_blank(),  # Remove major grid lines on Y-axis
    panel.grid.minor.y = element_blank(),  # Remove minor grid lines on Y-axis
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "right",  # Adjust legend position
    axis.title.x = element_blank()  # Remove the X-axis title
  ) +
  # Add dashed lines for the specified RLE thresholds
  geom_hline(yintercept = 0, linetype = "dashed", color = "green", size = 1) +  # 0% - Green
  geom_hline(yintercept = 30, linetype = "dashed", color = "yellow", size = 1) +  # 30% - Yellow
  geom_hline(yintercept = 50, linetype = "dashed", color = "orange", size = 1) +  # 50% - Orange
  geom_hline(yintercept = 80, linetype = "dashed", color = "red", size = 1) +  # 80% - Red
  # Add count of points (n) below the X-axis labels
  geom_text(data = ecoregion_counts, aes(x = ecoregion, y = -10, label = paste("n =", n)),
            size = 4, vjust = 1, color = "black") +  # Adjust position and text size
  guides(
    fill = "none"  # Remove legend for ecoregion fill
  ))

p1

ggsave("../../../figures/Benthic/HC_RS_ecoregion.png", plot = p1, width = 8, height = 6, dpi = 300)


```

```{r}
#| echo: false
# Preparing data for Visualization

# Summarizing the data
# Grouped calculation by ecoregion
viz_data_summary <- summarized %>%
  group_by(ecoregion) %>%
  summarise(
    # Identify maximum extent and corresponding severity range for each ecoregion
    max_extent = max(c(rel_sev_30, rel_sev_50, rel_sev_80), na.rm = TRUE),
    severity_range = case_when(
      max_extent == rel_sev_30 ~ "30-50",
      max_extent == rel_sev_50 ~ "50-80",
      max_extent == rel_sev_80 ~ "80-100",
      TRUE ~ NA_character_
    ),
    # Calculate max_severity from dat based on the identified severity range
    max_severity = case_when(
      severity_range == "30-50" ~ mean(
        dat %>% filter(ecoregion == first(ecoregion), 
                       relative_severity >= 30, 
                       relative_severity < 50) %>% pull(relative_severity),
        na.rm = TRUE
      ),
      severity_range == "50-80" ~ mean(
        dat %>% filter(ecoregion == first(ecoregion), 
                       relative_severity >= 50, 
                       relative_severity < 80) %>% pull(relative_severity),
        na.rm = TRUE
      ),
      severity_range == "80-100" ~ mean(
        dat %>% filter(ecoregion == first(ecoregion), 
                       relative_severity >= 80, 
                       relative_severity <= 100) %>% pull(relative_severity),
        na.rm = TRUE
      ),
      TRUE ~ NA_real_
    ),
    # Get status from the summarised dataframe
    status = first(status),
    .groups = "drop"
  )


# Ensure `status` is a factor for color mapping
viz_data_summary <- viz_data_summary %>%
  mutate(status = factor(status, levels = c("NT/LC", "VU", "EN", "CR", "CO")))

# Ensure shape_code is being applied to the plot
# Ensure ecoregion is a factor and map shape_code accordingly
viz_data_summary <- viz_data_summary %>%
  mutate(
    ecoregion = factor(ecoregion, levels = c("National", "Ngazidja", "Ndzuani", "Mwali")),
    shape_code = case_when(
      ecoregion == "National" ~ 16,  # Solid circle
      ecoregion == "Ndzuani" ~ 15,  # Solid square
      ecoregion == "Ngazidja" ~ 17,  # Solid triangle
      ecoregion == "Mwali" ~ 18,  # Solid diamond
      TRUE ~ NA_integer_
    )
  )

# Define IUCN RLE color coding
colors_class <- c(
  `CR` = "red",
  `EN` = "orange",
  `VU` = "yellow",
  `NT/LC` = "lawngreen"
)



viz_data_summary$ecoregion <- factor(viz_data_summary$ecoregion, levels = c("Ngazidja", "Ndzuani", "Mwali", "National"))


```

```{r}

# Load the shapefile of Comoros (replace with the correct path)
comoros_map <- st_read("../../../data_raw/spatial/ne_10m_admin_0_map_units.shp")

print(st_crs(comoros_map))

# Assign the CRS manually to EPSG:4326 (WGS 84)
comoros_map <- st_set_crs(comoros_map, 4326)

# Filter for the Comoros
comoros_data <- comoros_map[comoros_map$SOVEREIGNT == "Comoros", ]

st_crs(comoros_data)

str(dat)
dat$lat <- as.numeric(as.character(dat$lat))

dat_clean <- dat[!is.na(dat$lat) & !is.na(dat$long), ]

# Check the structure again to ensure missing values are handled
str(dat_clean)

dat_clean <- st_as_sf(dat_clean, coords = c("long", "lat"), crs = 4326)


# Map the IUCN status to colors
iucn_colors <- c(
  "CR" = "red",
  "EN" = "orange",
  "VU" = "yellow",
  "NT/LC" = "green"
)

# Extract longitude and latitude from the spatial object
dat_clean_coords <- st_coordinates(dat_clean)

# Add the extracted coordinates back to the data frame
dat_clean <- cbind(dat_clean, dat_clean_coords)

# Plot the data again
p_HC <- ggplot() +
  # Plot the filtered shapefile map of Comoros
  geom_sf(data = comoros_data, fill = "lightgray", color = "black") +
  # Plot points with reduced size
  geom_point(data = dat_clean, aes(x = X, y = Y, color = iucn_status), size = 2.5) +
  # Apply IUCN colors
  scale_color_manual(values = iucn_colors) +
  # Add title and labels
  labs(
    title = "",
    x = "Longitude",
    y = "Latitude",
    color = ""
  ) +
  # Adjust xlim and ylim for clarity
  coord_sf(xlim = c(43, 45), ylim = c(-12.7, -11), expand = FALSE) +
  # Minimal theme with reduced margins
  theme_minimal() +
  theme(
    legend.position = "bottom",       # Place the legend at the bottom
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), # Adjust margins for more space
    axis.title = element_text(size = 12),           # Increase axis label size
    plot.title = element_text(size = 16, hjust = 0.5) # Center and enlarge title
  )

p_HC

ggsave("../../../figures/Benthic/HC_RS_map.png", width = 12, height = 10, dpi = 300, bg = "white")


```

```{r, fig.cap= "Fig. 2; Criteria D results for Hard coral included Relative severity versus extent by ecoregion, color-coded by IUCN RLE status. Dashed lines represent threshold values for severity and extent (30%, 50%, 80%)."}
#| echo: false

# Create scatter plot with ecoregion names in the legend and shape codes in the plot
p2 <- ggplot(viz_data_summary, aes(x = max_severity, y = max_extent, shape = ecoregion, color = status)) +
  geom_point(size = 4, alpha = 0.7) +
  scale_x_continuous(
    limits = c(0, 100),
    name = "Relative Severity (%)",
    breaks = seq(0, 100, by = 20)
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    name = "Extent (%)",
    breaks = seq(0, 100, by = 20)
  ) +
  scale_shape_manual(
    values = c(16, 15, 17, 18),  # Shape codes directly mapped
    labels = c("National", "Ngazidja", "Ndzuani", "Mwali")  # Add labels for the legend
  ) +
  scale_color_manual(
    values = colors_class,
    name = "Status"
  ) +
  labs(
    title = "",
    x = "Relative Severity (%)",
    y = "Extent (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "blue", linetype = "dotted", size = 0.5),
    panel.grid.major.y = element_line(color = "green", linetype = "dotted", size = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  ) +
  # Add dashed lines for RLE thresholds
  geom_vline(xintercept = c(30, 50, 80), linetype = "dashed", color = "darkblue", size = 1) +
  geom_hline(yintercept = c(30, 50, 80), linetype = "dashed", color = "darkgreen", size = 1)

p2

ggsave("../../../figures/Benthic/HC_criteria_D_ecoregion.png", plot = p2, width = 8, height = 6, dpi = 300)

```

## Macroalgae Coral Ratio (MaCoR)

```{r}
#| echo: false
#| include: false
#| warning: false

# Set iteration levels
i_min <- 10
i_max <- 1e3

# Set iteration interval
i_interval <- 10

# Create empty object to hold results
criterion_d_biotic_disruption_MaCor_method_b <- tibble()

# Loop through iterations
for(i in seq(from = i_min, to = i_max, by = i_interval)) {

  # Set seed for reproducibility
  set.seed(i + 81)

  # Randomly assign baseline values
  dat <- current_macor_selected %>%
    mutate(
      baseline_cover = 0.2, # regional baseline
      baseline_sd = 0.1 # ragionel value
    ) %>%
    group_by(ecoregion) %>%
    mutate(
      baseline_cover = rnorm(1, mean = baseline_cover, sd = baseline_sd)
    )

  # Calculate relative severity
  dat <- dat %>%
    mutate(
      relative_severity = 100 * (baseline_cover - current_cover) / 
                          (baseline_cover - threshold_cover),
      relative_severity = scales::rescale(relative_severity, to = c(0, 100))
    )

  # Summarize proportion of stations for relative severity classes
  summarized <- dat %>%
    group_by(ecoregion) %>%
    summarise(
      rel_sev_30 = 100 * sum(relative_severity >= 30 & relative_severity < 50) / length(relative_severity),
      rel_sev_50 = 100 * sum(relative_severity >= 50 & relative_severity < 80) / length(relative_severity),
      rel_sev_80 = 100 * sum(relative_severity >= 80 & relative_severity <= 100) / length(relative_severity),
      relative_severity = mean(relative_severity, na.rm = TRUE), # Add mean relative severity
      .groups = "drop"
    )

  # Assign threat status
  summarized <- summarized %>%
    mutate(
      status_30 = ifelse(rel_sev_30 >= 80 & rel_sev_30 <= 100, 2, NA),
      status_50 = ifelse(rel_sev_50 >= 80 & rel_sev_50 <= 100, 3, NA),
      status_50 = ifelse(rel_sev_50 >= 50 & rel_sev_50 < 80, 2, status_50),
      status_80 = ifelse(rel_sev_80 >= 50 & rel_sev_80 < 80, 3, NA),
      status_80 = ifelse(rel_sev_80 >= 80 & rel_sev_80 <= 100, 4, status_80),
      status_80 = ifelse(rel_sev_80 >= 30 & rel_sev_80 < 50, 2, status_80),
      status_30 = ifelse(is.na(status_30), 1, status_30),
      status_50 = ifelse(is.na(status_50), 1, status_50),
      status_80 = ifelse(is.na(status_80), 1, status_80),
      max_threat = pmax(status_30, status_50, status_80)
    )

  # Create conversion object for threat values
  threat_conversions <- tribble(
    ~threat_value, ~status,
    1, "NT/LC",
    2, "VU",
    3, "EN",
    4, "CR",
    5, "CO"
  )

  # Convert threat values
  summarized <- summarized %>%
    left_join(threat_conversions %>% rename(max_threat = threat_value), by = "max_threat")

  # Include iteration number
  summarized <- summarized %>%
    mutate(Iteration = i)

  # Harvest results
  criterion_d_biotic_disruption_MaCor_method_b <- bind_rows(
    criterion_d_biotic_disruption_MaCor_method_b,
    summarized
  )
}

    
    
# summarise
    criterion_d_biotic_disruption_MaCor_method_b %>%
      group_by(ecoregion,
               status) %>%
      summarise(n_categories = n()) %>%
      mutate(percent = 100 * n_categories / sum(n_categories))              
```

```{r}
#| echo: false
#| include: false

# Create a new column for IUCN RLE categories
dat <- dat %>%
  mutate(
    iucn_status = case_when(
      relative_severity < 30 ~ "NT/LC",  # Green
      relative_severity >= 30 & relative_severity < 50 ~ "VU",  # Yellow
      relative_severity >= 50 & relative_severity < 80 ~ "EN",  # Orange
      relative_severity >= 80 & relative_severity <= 100 ~ "CR",  # Red
      TRUE ~ NA_character_
    )
  )

# Define the IUCN RLE color coding
iucn_colors <- c(
  "CR" = "red",
  "EN" = "orange",
  "VU" = "yellow",
  "NT/LC" = "green"
)

# Calculate the number of points for each ecoregion
ecoregion_counts <- dat %>%
  group_by(ecoregion) %>%
  summarise(n = n(), .groups = "drop")

dat <- dat %>%
  mutate(iucn_status = factor(iucn_status, levels = c("NT/LC", "VU", "EN", "CR"))) 


dat_mcor <- dat


```

```{r, fig.cap= "Fig. 1; Relative severity for MaCoR by ecoregion, with points color-coded according to IUCN RLE status (NT/LC, VU, EN, CR). The number of data points for each ecoregion is indicated by n"}
#| echo: false


# Create the plot
p3 <- suppressWarnings(
  ggplot(dat, aes(x = ecoregion, y = relative_severity)) +
  geom_boxplot(aes(fill = ecoregion), alpha = 0.5, outlier.shape = NA) +  # Boxplot with different fill colors for ecoregions, remove outliers from boxplot
  geom_point(aes(color = iucn_status), alpha = 0.7, size = 3, position = position_jitter(width = 0.2)) +  # Add scattered points with jitter
  scale_y_reverse(  # Reverse the Y-axis
    name = "Relative Severity",
    breaks = seq(0, 100, by = 20)
  ) +
  scale_color_manual(
    values = iucn_colors,  # Apply color mapping based on IUCN status
    name = "Relative Severity"
  ) +
  labs(
    title = "",
    x = "",
    y = "Relative Severity (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "blue", linetype = "dotted", size = 0.5),
    panel.grid.major.y = element_blank(),  # Remove major grid lines on Y-axis
    panel.grid.minor.y = element_blank(),  # Remove minor grid lines on Y-axis
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    legend.position = "right",  # Adjust legend position
    axis.title.x = element_blank()  # Remove the X-axis title
  ) +
  # Add dashed lines for the specified RLE thresholds
  geom_hline(yintercept = 0, linetype = "dashed", color = "green", size = 1) +  # 0% - Green
  geom_hline(yintercept = 30, linetype = "dashed", color = "yellow", size = 1) +  # 30% - Yellow
  geom_hline(yintercept = 50, linetype = "dashed", color = "orange", size = 1) +  # 50% - Orange
  geom_hline(yintercept = 80, linetype = "dashed", color = "red", size = 1) +  # 80% - Red
  # Add count of points (n) below the X-axis labels
  geom_text(data = ecoregion_counts, aes(x = ecoregion, y = -10, label = paste("n =", n)),
            size = 4, vjust = 1, color = "black") +  # Adjust position and text size
  guides(
    fill = "none"  # Remove legend for ecoregion fill
  )) 

p3

ggsave("../../../figures/Benthic/MaCoR_RS_ecoregion.png", plot = p3, width = 8, height = 6, dpi = 300)

```

```{r}

str(dat)
dat$lat <- as.numeric(as.character(dat$lat))

dat_clean <- dat[!is.na(dat$lat) & !is.na(dat$long), ]

# Check the structure again to ensure missing values are handled
str(dat_clean)

dat_clean <- st_as_sf(dat_clean, coords = c("long", "lat"), crs = 4326)


# Map the IUCN status to colors
iucn_colors <- c(
  "CR" = "red",
  "EN" = "orange",
  "VU" = "yellow",
  "NT/LC" = "green"
)

# Extract longitude and latitude from the spatial object
dat_clean_coords <- st_coordinates(dat_clean)

# Add the extracted coordinates back to the data frame
dat_clean <- cbind(dat_clean, dat_clean_coords)

# Plot the data again
p_MaCoR <- ggplot() +
  # Plot the filtered shapefile map of Comoros
  geom_sf(data = comoros_data, fill = "lightgray", color = "black") +
  # Plot points with reduced size
  geom_point(data = dat_clean, aes(x = X, y = Y, color = iucn_status), size = 2.5) +
  # Apply IUCN colors
  scale_color_manual(values = iucn_colors) +
  # Add title and labels
  labs(
    title = "",
    x = "Longitude",
    y = "Latitude",
    color = ""
  ) +
  # Adjust xlim and ylim for clarity
  coord_sf(xlim = c(43, 45), ylim = c(-12.7, -11), expand = FALSE) +
  # Minimal theme with reduced margins
  theme_minimal() +
  theme(
    legend.position = "bottom",       # Place the legend at the bottom
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), # Adjust margins for more space
    axis.title = element_text(size = 12),           # Increase axis label size
    plot.title = element_text(size = 16, hjust = 0.5) # Center and enlarge title
  )

p_MaCoR

ggsave("../../../figures/Benthic/MaCoR_RS_map.png", width = 12, height = 10, dpi = 300, bg = "white")
```

```{r}
#| echo: false
# Preparing data for Visualization

# Summarizing the data
# Grouped calculation by ecoregion
viz_data_summary <- summarized %>%
  group_by(ecoregion) %>%
  summarise(
    # Identify maximum extent and corresponding severity range for each ecoregion
    max_extent = max(c(rel_sev_30, rel_sev_50, rel_sev_80), na.rm = TRUE),
    severity_range = case_when(
      max_extent == rel_sev_30 ~ "30-50",
      max_extent == rel_sev_50 ~ "50-80",
      max_extent == rel_sev_80 ~ "80-100",
      TRUE ~ NA_character_
    ),
    # Calculate max_severity from dat based on the identified severity range
    max_severity = case_when(
      severity_range == "30-50" ~ mean(
        dat %>% filter(ecoregion == first(ecoregion), 
                       relative_severity >= 30, 
                       relative_severity < 50) %>% pull(relative_severity),
        na.rm = TRUE
      ),
      severity_range == "50-80" ~ mean(
        dat %>% filter(ecoregion == first(ecoregion), 
                       relative_severity >= 50, 
                       relative_severity < 80) %>% pull(relative_severity),
        na.rm = TRUE
      ),
      severity_range == "80-100" ~ mean(
        dat %>% filter(ecoregion == first(ecoregion), 
                       relative_severity >= 80, 
                       relative_severity <= 100) %>% pull(relative_severity),
        na.rm = TRUE
      ),
      TRUE ~ NA_real_
    ),
    # Get status from the summarised dataframe
    status = first(status),
    .groups = "drop"
  )



# Ensure `status` is a factor for color mapping
viz_data_summary <- viz_data_summary %>%
  mutate(status = factor(status, levels = c("NT/LC", "VU", "EN", "CR", "CO")))

# Ensure shape_code is being applied to the plot
# Ensure ecoregion is a factor and map shape_code accordingly
viz_data_summary <- viz_data_summary %>%
  mutate(
    ecoregion = factor(ecoregion, levels = c("National", "Ngazidja", "Ndzuani", "Mwali")),
    shape_code = case_when(
      ecoregion == "National" ~ 16,  # Solid circle
      ecoregion == "Ndzuani" ~ 15,  # Solid square
      ecoregion == "Ngazidja" ~ 17,  # Solid triangle
      ecoregion == "Mwali" ~ 18,  # Solid diamond
      TRUE ~ NA_integer_
    )
  )

# Define IUCN RLE color coding
colors_class <- c(
  `CR` = "red",
  `EN` = "orange",
  `VU` = "yellow",
  `NT/LC` = "lawngreen"
) 

```

```{r, fig.cap= "Fig. 2; Criteria D results for MaCor included Relative severity versus extent by ecoregion, color-coded by IUCN RLE status. Dashed lines represent threshold values for severity and extent (30%, 50%, 80%)."}
#| echo: false

# Create scatter plot with ecoregion names in the legend and shape codes in the plot
p4 <- ggplot(viz_data_summary, aes(x = max_severity, y = max_extent, shape = ecoregion, color = status)) +
  geom_point(size = 4, alpha = 0.7) +
  scale_x_continuous(
    limits = c(0, 100),
    name = "Relative Severity (%)",
    breaks = seq(0, 100, by = 20)
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    name = "Extent (%)",
    breaks = seq(0, 100, by = 20)
  ) +
  scale_shape_manual(
    values = c(16, 15, 17, 18),  # Shape codes directly mapped
    labels = c("National", "Ngazidja", "Ndzuani", "Mwali")  # Add labels for the legend
  ) +
  scale_color_manual(
    values = colors_class,
    name = "Status"
  ) +
  labs(
    title = "",
    x = "Relative Severity (%)",
    y = "Extent (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_line(color = "blue", linetype = "dotted", size = 0.5),
    panel.grid.major.y = element_line(color = "green", linetype = "dotted", size = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(color = "black")
  ) +
  # Add dashed lines for RLE thresholds
  geom_vline(xintercept = c(30, 50, 80), linetype = "dashed", color = "darkblue", size = 1) +
  geom_hline(yintercept = c(30, 50, 80), linetype = "dashed", color = "darkgreen", size = 1)  

p4

ggsave("../../../figures/Benthic/MaCoR_criteria_D_ecoregion.png", plot = p4, width = 8, height = 6, dpi = 300)

```

# Saving csv file - Sites RS values

```{r}


# RS values for sites

dat_hc$indicator <- "HC"
dat_mcor$indicator <- "MaCoR"

# merge
dat_RS <- rbind(dat_hc, dat_mcor)

# Order columns

dat_RS <- dat_RS[, c("ecoregion", "Site", "lat", "long","indicator", "relative_severity", "iucn_status")]

dat_RS <- dat_RS[dat_RS$ecoregion != "National", ]


write.csv(dat_RS, "../../../data/RS_values_site_benthic.csv")



```
